---
title: "neoantigen statistical tests"
output: word_document
---


```{r}
#libraries
library(tidyverse)
library(rstatix) #for tests with effect sizes
library(ggpubr) #plots with annotations

#load data
data <- read_csv("../../processed/final_predicted.csv")
head(data)
```
#create immunogenicity groups
```{r}
#create group labels
data$status <- ifelse(
  data$label == 1, "immunogenic", "non-immunogenic"
)

#set order
data$status <- factor(
  data$status, 
  levels=c("non-immunogenic", "immunogenic")
)
table(data$status)
```


# CONFOUND CHECKS
#if some factors are affecting results
```{r}
#peptide length check

#ensure different length dist do not cause differences
length_table <- table(data$mt_length, data$status)
print(length_table)

#is length distribution independent of status? 
length_chi <- chisq.test(length_table)
print(length_chi)
```
- p value < 0.05 length IS associated with immunogenicity
- likely confounds results
- maybe stratify by length as well 

```{r}
#effect size for association 
length_v <- cramer_v(length_table)
print(length_v)
```
Not very strong

```{r}
#plot values
ggplot(data, aes(x = factor(mt_length), fill=status)) + geom_bar(position="fill") + labs(title = "proportion immunogenic by length", x="peptide length", y = "proportion")
```

#HLA check 
```{r}
#same logic for hla distribution
hla_table <- table(data$best_hla, data$status)
print(hla_table)
```
#many HLAs would add noise
#keep only most common
```{r}
top_hlas <- data %>% count(best_hla, sort=TRUE) %>% head(10) %>% pull(best_hla)
data$hla_group <- ifelse(data$best_hla %in% top_hlas, data$best_hla, "other")

#again chi sq test
hla_table <- table(data$hla_group, data$status)
print(hla_table)

hla_chi <- chisq.test(hla_table)
print(hla_chi)
```
ALSO CONFOUNDS (wow)

#but how severe confounding
```{r}
hla_v <- cramer_v(hla_table)
print(hla_v)
```
Association bw 0.1-0.3 so weak. 


#confound effect sizes chart 
```{r}
#make df for plot
confound_df <- data.frame(factor=c("peptide length", "HLA allele"), 
                          cramers_v = c(length_v, hla_v))
head(confound_df)
```
```{r}
#barshart
p1 <- ggplot(confound_df, aes(x=cramers_v, y=factor)) + geom_col(fill = "#2c5981", width = 0.5) + geom_text(aes(label=sprintf("V = %.3f", cramers_v)), hjust=-0.15) + scale_x_continuous(limits=c(0,0.5)) + labs(title="Confound check", x="cramers v", y="") + theme_pubr()
print(p1)
```



#PAIRED WT-MT ANALYSIS
does mutation change binding affinity? 
```{r}
#paired
#mt #wt to same peptide #length and HLA same 
#log since affinities are skewed 

#long format for easier 
long <- data %>% 
  select(epitope_id, status, wt_affinity_percentile, mt_affinity_percentile) %>%
  pivot_longer(cols = c(wt_affinity_percentile, mt_affinity_percentile), names_to="type",
    values_to="affinity_percentile")

#clean labels
long$type[long$type=="wt_affinity_percentile"] <- "WT"
long$type[long$type=="mt_affinity_percentile"] <- "MT"

#set order
long$type <- factor(long$type, levels=c("WT", "MT"))
head(long)
```

```{r}
summary(long$affinity_percentile)
```

```{r}
#constant to remove zeroes
#log transform
long$neg_log10 <- -log10(long$affinity_percentile + 0.0001) #reverses axis #higher = stronger binding
head(long)
```


#PAIRED WILCOXON TEST
#do mutations change binding? 
```{r}
#affinity values skewed #use log
#lower = stronger
paired_test <- long %>% group_by(status) %>% wilcox_test(neg_log10 ~ type, paired=TRUE) %>% add_significance()
print(paired_test)
```
- both values < 0.05
- mutation consistently changes binding both both groups
- but how much? 

```{r}
#effect size 
#large dataset can have small p-values
paired_eff <- long %>% group_by(status) %>% wilcox_effsize(neg_log10 ~ type, paired=TRUE)
print(paired_eff)
```
```{r}
#violin plot
ggviolin(long, x="type", y= "neg_log10", group = "epitope_id", fill="type", add="boxplot") + facet_wrap(~status) + stat_compare_means(method="wilcox.test", label="p.format") + theme_pubr()
```

```{r}
#plot
ggplot(long, aes(x=type, y=neg_log10, group=epitope_id, color=status)) + geom_line(alpha=0.05) + facet_wrap(~status) 
```


#BW GROUPS
Do immunogenic peptides show different binding changes?

#log2 fold change (log2(mt/wt))
```{r}
data$log2fc <- log2((data$mt_affinity_percentile + 0.0001) / (data$wt_affinity_percentile + 0.0001)) 
summary(data$log2fc)
```

#mann-whitney 
```{r}
#do immunogenic peptides have better binding change? 
mw_test <- wilcox.test(log2fc ~ status, data=data)
mw_effect <- wilcox_effsize(log2fc ~ status, data=data)
print(mw_test)
print(mw_effect)
```
```{r}
#make summary table
summary_df <- data.frame(
  test = c("paired mt vs wt nonimmunogenic", 
           "paired mt vs wt immunogenic", 
           "between groups log2fc"), 
  effsize = c(paired_eff$effsize[1], paired_eff$effsize[2], mw_effect$effsize[1])
)
head(summary_df)
```

```{r}
#plot graph
ggplot(summary_df, aes(x=effsize, y=fct_rev(factor(test)))) + geom_point(size=4) + geom_text(aes(label=sprintf("r=%.3f", effsize)), vjust=-0.5) + xlim(-0.01, 0.7)+ labs(x = "rank r", y="") + theme_pubr()
```


```{r}
#boxplot
ggboxplot(data, x="status", y="log2fc", fill="status") + stat_compare_means(method="wilcox.test", label="p.format") 
```

#stratified by length
```{r}
ggplot(data, aes(x=log2fc, fill=status)) + geom_density(alpha=0.5) + facet_wrap(~mt_length, nrow=1, labeller = label_both) +theme_pubr()
```

#DIRECTION OF CHANGE
- did mutation strength/weaken binding?
```{r}
data$direction <- "unchanged"
data$direction[data$mt_affinity_percentile < data$wt_affinity_percentile] <- "stronger"
data$direction[data$mt_affinity_percentile > data$wt_affinity_percentile] <- "weaker"
table(data$direction)
```

#fisher's exact test 
are immunogenic peptides more likely to strengthen binding? 
```{r}
data$stronger <- data$direction == "stronger"
fisher_stronger <- fisher.test(table(data$status, data$stronger))
print(fisher_stronger)
```
#gain and loss of presentation
- did mut create/destroy presentation? 
```{r}
data$gained_pres <- data$wt_presented == 0 & data$mt_presented == 1
data$lost_pres <- data$wt_presented == 1 & data$mt_presented == 0
head(data)
```
```{r}
table(data$gained_pres)
```
```{r}
table(data$lost_pres)
```

#fisher's exact for gain 
```{r}
fisher_gain <- fisher.test(table(data$status, data$gained_pres))
print(fisher_gain)
```
#fisher's exact for loss 
```{r}
fisher_loss <- fisher.test(table(data$status, data$lost_pres))
print(fisher_loss)
```
#HLA distribution of dataset
```{r}
hla_freq <- data %>% count(best_hla, sort=TRUE) %>% slice_max(n, n=10) 

ggplot(hla_freq, aes(n, reorder(best_hla, n))) + geom_col(fill="grey") + geom_text(aes(label=n)) + labs(x="count", y="") + theme_pubr()
```


#subgroup composition 
```{r}
data$presented <- ifelse(data$mt_presented == 1, "presented", "not presented")
subgroup_count <- data %>% count(presented, status) %>% mutate(group = paste(presented, status, sep="_"))
head(subgroup_count)
```
```{r}
ggplot(subgroup_count, aes(x=n, y=group, fill=status)) + geom_col() + geom_text(aes(label=n)) + theme_pubr()
```


#logistic regression 
do binding features predict immunogenicity? 
```{r}
#turn length into categories
data$length_factor <- factor(data$mt_length)
table(data$length_factor)
```
```{r}
#binding features only
model <- glm(label ~ log2fc + mt_affinity_percentile + wt_affinity_percentile, data=data, family=binomial)
summary(model)
```

#also control for peptide length/HLA
```{r}
control_model <- glm(label ~ log2fc + mt_affinity_percentile + wt_affinity_percentile + length_factor + hla_group, data=data, family=binomial)
summary(control_model)
```

#checking if within group holds if separated by length
```{r}
#stratified by length
strat_test <- data %>% group_by(mt_length) %>% wilcox_test(log2fc~status)
print(strat_test)
```
```{r}
strat_eff <- data %>% group_by(mt_length) %>% wilcox_effsize(log2fc~status)
print(strat_eff)
```

```{r}
#likelihood ratio 
anova(model, control_model, test="Chisq")
```
- confirms controlled model is significantly better



