---
title: "mag stats"
output: word_document
---

- L1 magnitude is mt-wt 
- Does size of mutation shift differ between pairs where mutations flipped to immunogenic vs not? 

```{r}
#load packages
library(tidyverse)
library(rstatix)
library(ggpubr)
```

```{r}
#load 
df <- read_csv("../../output/final_embeds/analysis_df.csv")
head(df)
```

```{r}
#make label ordered factor
df$label <- ifelse(df$label == 1, "immunogenic", "non-immunogenic")
df$label <- factor(df$label,
                        levels = c("non-immunogenic", "immunogenic"))
df$model <- factor(df$model, levels = c("ESM-2", "ProtTrans"))
head(df)
```

```{r}
#rows per model
table(df$model)
```

```{r}
#reshape to long for analysis
#prottrans has no bos, so abs_sum_bos can be dropped 

long <- df %>% pivot_longer(cols=c(abs_sum_bos, abs_sum_eos, abs_sum_mutpos, mean_abs_allpos), names_to="criteria", values_to="value") %>% filter(!is.na(value))
head(long)
```
```{r}
#over all test
#mann-whitney u #unpaired #either flipped or not
overall_p <- long %>% group_by(model, criteria) %>% wilcox_test(value ~ label) %>% adjust_pvalue(method="BH") %>% ungroup()
print(overall_p)
```
- BOS esm-2 has small significance (padj = 0.042 < 0.05)
- does seem to have small effect on whether it flips or not 

```{r}
#effect sizes 
overall_r <- long %>% group_by(model, criteria) %>% wilcox_effsize(value ~ label) %>% ungroup()
print(overall_r)
```
- effsize = 0.035 so small magnitude 


```{r}
#nicer labels
plot_data <- left_join(overall_r, overall_p,
                       by = c("model","criteria"))

plot_data$criteria <- recode(plot_data$criteria, 
                             "abs_sum_bos" = "BOS token",
                             "abs_sum_eos" = "EOS token", 
                             "abs_sum_mutpos" = "Mutated residue", 
                             "mean_abs_allpos" = "Mean positions")

ggplot(plot_data, aes(x = effsize, y = criteria, shape = model, color = p.adj < 0.05)) + geom_point(size = 4, position = position_dodge(width = 0.5)) + scale_color_manual(values = c("FALSE" = "salmon", "TRUE" = "steelblue"), labels = c("n.s.", "p.adj < 0.05")) + labs(x = "Effect size (r)", y = NULL, color = NULL, shape = "Model") + theme_classic()
```


```{r}
#length stratified test
strat_p <- long %>% group_by(model, criteria, mt_length) %>% wilcox_test(value ~ label) %>% adjust_pvalue(method = "BH") %>% ungroup()
print(strat_p, n=30)
```
- sig for many disappears when p adjusted
- is p adj too harsh? many n values do inflate p values 

```{r}
strat_r <- long %>% group_by(model, criteria, mt_length) %>% wilcox_effsize(value ~ label) %>% ungroup()
print(strat_r, n=30)
```

```{r}
#heatmap of length stratified 
heat_data <- left_join(
  strat_r,
  strat_p,
  by = c("model","criteria","mt_length")
)

ggplot(heat_data, aes(x = factor(mt_length), y=criteria, fill = p)) + geom_tile(color="white") + geom_text(aes(label=round(effsize,2)), size=3) + facet_wrap(~model) + scale_fill_gradient(low="white", high="steelblue") + labs(x="peptide length", y = NULL, fill="r") + theme_minimal()
```



#distribution of absolute sum values for supplementary information 
#split imm/nonimm
```{r}
#is mutation shift higher when wt turns immunogenic 
#two proportion test #loop over both

#split by label
df_non <- df %>% filter(label == "non-immunogenic")
df_imm <- df %>% filter(label == "immunogenic")

#summarize
sum_non <- df_non %>% group_by(model) %>% summarise(s_non=sum(top1_correct), n_non=n(), prop_non = s_non/n_non)

sum_imm <- df_imm %>% group_by(model) %>% summarise(s_imm=sum(top1_correct), n_imm=n(), prop_imm = s_imm/n_imm)

#join tables
signal <- sum_non %>% inner_join(sum_imm, by="model")
print(signal)
```
```{r}
#compute pvalues
signal$p_value <- mapply(
  function(s_non, s_imm, n_non, n_imm) {
    prop.test(c(s_non, s_imm), c(n_non, n_imm))$p.value
  }, 
  signal$s_non, 
  signal$s_imm, 
  signal$n_non, 
  signal$n_imm
)

head(signal)
```
- mutation strength shift not higher when peptide turns immunogenic 


```{r, fig.width = 10, fig.height = 8}
library(tidyverse)
library(ggpubr)
library(rstatix)

# 1. Clean up the labels
long <- long %>%
  mutate(criteria = recode(criteria,
    abs_sum_bos = "BOS token",
    abs_sum_eos = "EOS token",
    abs_sum_mutpos = "Mutated residue",
    mean_abs_allpos = "Mean over residues"
  ))

# 2. Run statistics and automatically add label positions
stat.test <- long %>%
  group_by(model, criteria) %>%
  wilcox_test(value ~ label) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj") %>%
  add_y_position(formula = value ~ label, scales = "free_y") # Simplified here

# 3. Create the plot
ggplot(long, aes(x = label, y = value, fill = label)) +
  geom_violin() +
  geom_boxplot(width = 0.18) +
  
  # Add p-value stars
  stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0.01) +
  
  # Grid layout
  facet_grid(model ~ criteria, scales = "free_y") +
  
  # Custom colors and styling
  labs(x = "", y = "L1 embedding shift") +
  theme_pubclean() + theme(axis.text.x = element_text(angle=35, hjust=1)) 

```





