---
title: "R Notebook"
output: word_document
---

```{r}
#load packages
library(tidyverse)
library(rstatix)
library(ggpubr)
```

```{r}
#load 
df <- read_csv("../../output/final_embeds/analysis_df.csv")
head(df)
```
```{r}
#make label ordered factor
df$label <- ifelse(df$label == 1, "immunogenic", "non-immunogenic")
df$label <- factor(df$label,
                        levels = c("non-immunogenic", "immunogenic"))
df$model <- factor(df$model, levels = c("ESM-2", "ProtTrans"))
head(df)
```

### POSITION ANALYSIS
```{r}
#does position affect immunogenicity?
print(table(df$mut_pos))
```

```{r}
pos_table <- table(df$label, df$mut_pos)
print(pos_table)
```
#chisq test
```{r}
#does mutpos differ between imm/nonimm? 
chisq_pos <- chisq.test(pos_table)
print(chisq_pos)
```
- position is not random wt imm  

```{r}
#position strat analysis
#compare each pos separately
pos_analysis <- df %>% group_by(model, mut_pos) %>% 
  summarise(n_total = n(), n_imm = sum(label=="immunogenic"), prop_imm = n_imm/n_total, .groups="drop") %>% arrange(model, mut_pos)
print(pos_analysis, n=30)
```
- 4-5 is 0.25
- 6 is 0.34 highest of all = most immunogenic 
- 10-11 is 0.25 to 0.15 == maybe due to length confound (exists only in longest pep)

```{r}
#reshape to long for analysis
#prottrans has no bos, so abs_sum_bos can be dropped 

long <- df %>% pivot_longer(cols=c(abs_sum_bos, abs_sum_eos, abs_sum_mutpos, mean_abs_allpos), names_to="criteria", values_to="value") %>% filter(!is.na(value))
head(long)
```

#wilcox test
#not independent
```{r}
#test each pos for non/imm difference
pos_wilcox <- long %>% group_by(model, criteria, mut_pos) %>% wilcox_test(value ~ label) %>% adjust_pvalue(method="BH") %>% ungroup()
print(pos_wilcox, n=100)
```
so strength does NOT differ between imm/nonimm at any specific position 

#logistic regression
```{r}
#does position predict imm? 
#test for each model differently 

df_esm2 <- df %>% filter(model == "ESM-2")
  
#change to numeric for logreg
df_esm2$imm_numeric <- as.numeric(df_esm2$label) - 1 #0 nonimm, 1=imm
  
#fit logreg
glm_esm2 <- glm(imm_numeric ~ mut_pos, data=df_esm2, family="binomial")
print(summary(glm_esm2))
```
```{r}
#calculate odds ratio 
odds_ratio_esm2 <- exp(coef(glm_esm2)["mut_pos"])
print(round(odds_ratio_esm2),4)
```

```{r}
#does position predict imm? 
#test for each model differently 

df_prot <- df %>% filter(model == "ProtTrans")
  
#change to numeric for logreg
df_prot$imm_numeric <- as.numeric(df_prot$label) - 1 #0 nonimm, 1=imm
  
#fit logreg
glm_prot <- glm(imm_numeric ~ mut_pos, data=df_prot, family="binomial")
print(summary(glm_prot))
```
```{r}
#calculate odds ratio 
odds_ratio_prot <- exp(coef(glm_prot)["mut_pos"])
print(round(odds_ratio_prot),4)
```

- almost identical logreg output 
- stat sig but not large effect
position explains very little imm 

#GRAPHS
```{r}
#prop imm by position 
pos_prop <- df %>% group_by(mut_pos) %>% summarise(
  n_total = n() / 2, n_imm = sum(label=="immunogenic") /2, 
  prop_imm = n_imm/n_total, 
  se = sqrt(prop_imm * (1-prop_imm) / n_total)
)
ggplot(pos_prop, aes(x = factor(mut_pos), y = prop_imm)) + geom_col(fill = "steelblue") + geom_errorbar(aes(ymin = prop_imm - se, ymax = prop_imm + se)) + geom_hline(yintercept = mean(pos_prop$prop_imm), linetype = "dashed", color = "grey") + labs(x = "Mutation position", y = "Proportion immunogenic") + theme_classic()
```

```{r}
#imm/nonimm by pos
pos_count <- df %>% filter(model == "ESM-2") %>%   # drop one model since counts are identical
  group_by(mut_pos, label) %>% summarise(count = n(), .groups = "drop")

ggplot(pos_count, aes(x = factor(mut_pos), y = count, fill = label)) + geom_col() + labs(x = "mutation position", y = "count", fill = NULL) + theme_classic() 
```

#length 11 has very few immunogenic
#likely fewer long peptides
#control for length
```{r}
glm_length_esm2 <- glm(imm_numeric ~ mut_pos + mt_length, data=df_esm2, family="binomial")
summary(glm_length_esm2)
```

```{r}
glm_length_prot <- glm(imm_numeric ~ mut_pos + mt_length, data=df_prot, family="binomial")
summary(glm_length_prot)
```
LENGTH MORE SIGNIFICANT, CONFOUNDED PREVIOUS RESULT 

```{r}
#does interaction add anything?
anova_esm2 <- anova(glm_esm2, glm_length_esm2, test="Chisq")
print(anova_esm2)
```
```{r}
anova_prot <- anova(glm_prot, glm_length_prot, test="Chisq")
print(anova_prot)
```
YES EXPLAINS MUCH MORE THAN OTHER MODEL

#treat position as categorical with length
```{r}
#esm2 first
df_esm2$mut_pos_f <- factor(df_esm2$mut_pos)
glm_cat_esm2 <- glm(imm_numeric ~ mut_pos_f + mt_length, data=df_esm2, family="binomial")
summary(glm_cat_esm2)
```


```{r}
#esm2 first
df_prot$mut_pos_f <- factor(df_prot$mut_pos)
glm_cat_prot <- glm(imm_numeric ~ mut_pos_f + mt_length, data=df_prot, family="binomial")
summary(glm_cat_prot)
```
POS 4 AND 5 LOWER ODDS OF IMM 
POS 6 HAS SLIGHTLY HIGHER ODDS


#WHICH POS DRIVES CHISQ 
```{r}
#test one pos vs all others combined
pos <- sort(unique(df$mut_pos))

#results table
results <- data.frame(mut_pos=pos, p_fisher=NA_real_)

#loop over pos
for (i in seq_along(pos)) {
  p <- pos[i]
  #is it pos
  is_pos <- df$mut_pos == p
  tab <- table(df$label, is_pos) 
  results$p_fisher[i] <- fisher.test(tab)$p.value
}

#adj for comparision
results$p_adj <- p.adjust(results$p_fisher, method="BH")
print(results)
```
#fisher's by position
```{r}
ggplot(results, aes(x = factor(mut_pos), y = -log10(p_adj))) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
  labs(x = "Mutation position", y = "-log10(adjusted p-value)",
       caption = "Dashed line = p.adj 0.05") +
  theme_classic()
```

```{r}
#wilcox pvalue heatmap across pos and features
pos_wilcox_clean <- pos_wilcox %>% mutate(criteria = recode(criteria, 
                                                            "abs_sum_bos" = "BOS token", "abs_sum_eos" = "EOS token", "abs_sum_mutpos" = "Mutated residue", "mean_abs_allpos" = "Mean positions"))

#heatmap of length stratified 
ggplot(pos_wilcox_clean, aes(x = factor(mut_pos), y=criteria, fill = p.adj)) + geom_tile(color="white") + geom_text(aes(label=round(p.adj, 2)), size=2) + facet_wrap(~model) + scale_fill_gradient(low="white", high="steelblue") + labs(x="mutation position", y = NULL, fill="p.adj") + theme_minimal()
```

